Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-09 draft.tex
Update notes

\begin{abstract}
  Our goal is to obtain a complete set of angular observables
  arising in a generic multi-body process without the need to carry
  out a likelihood fit of the angular distribution to the measured
  events. Instead, we apply the method of moments that relies
  both on the orthogonality of angular functions and the estimation of
  integrals by Monte Carlo techniques. The main advantage of
  this method is that the joint distribution of all observables
  given the data can be easily extracted, even for very few
  events. Our main contribution is to show explicitly how systematic
  uncertainties from detector-resolution and acceptance effects can
  be incorporated in a general algorithm. Finally, we present the
  necessary process-dependent formulae needed for direct
  implementation of the method for several rare decays of
  interest.
\end{abstract}

@@

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare decay FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, note that the method we
describe in the following is general, and it applies to arbitrary decay
or scattering processes. We find a previous work \cite{Dighe:1998vk}
that advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays, but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this previous
work by studying the uncertainties that are introduced by mismodelling
of the angular distribution, and by working out a recipe to determine
and unfold detector effects. Moreover, we show
that the method of moments has \add{several} major advantages over the usual
approach based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrization and approximations.

    \item \add{Likelihood fits of problems, for which the underlying
        physical model is only partially known, may be delicate. Such cases can lead to
        overestimating the number of physical parameters, and consequently
        inhibit the convergence of the fits. (As an example, this type
        of problem occured in toy studies of the decay $B\to K^*(\to K\pi)\ell^+\ell^-$
        as reported in \cite{Egede:1142152}. It was subsequently solved
        when a missing symmetry relation between the angular observables
        was found and applied, thereby reducing the number of fit parameters.)
    }
    \item Mismodeling the underlying physics model can result in systematic bias
        in likelihood fits. We will
        show that the method of moments is insensitive to a certain type of mismodeling;
        i.e. introducing a cutoff in a partial-wave expansion of the signal PDF.

@@

We continue with basic definitions that pertain to angular observables, and our
results in the subsequent sections. Let $\thvec$ denote the set of all
angles, and let $\nuvec$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $S_i$ as a coefficient in the probability density
function (PDF) of the process by means of
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) \equiv \sum_i S_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$\{f_i(\thvec)\}$. We assume there exists a dual basis of functions
$\{\dual{f}_i(\thvec)\}$ such that the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega \rmdx{\vec{\theta}} \dual{f}_i(\thvec) f_j(\thvec)  = \delta_{ij}
\end{equation}
hold with $\Omega$ representing the full angular phase space relevant
to the process.  For particle decays, $P$ is generally expressed in
terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\rmdx{^{2}\Gamma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\rmdx{^{2}\sigma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process. This change of normalization is equivalent to a linear rescaling of
the angular observables $\lbrace S_i\rbrace$; thus ratios or similar suitable combinations
of the angular observables are not affected by this change of normalization.\\

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. It turns out to be much more convenient to work
in the basis of Legendre polynomials but the statistical properties
are equivalent for both approaches provided that one replace $f_i \to
\dual{f}_i$ appropriately. Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:lambdabtolambdall}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

In the remainder of this letter we discuss how to obtain an angular
observable $S_i(\nuvec)$ in an experimental setup where each recorded
event is (approximately) distributed according to $P$.  We establish
the statistical basics in section \ref{sec:sample-based-det}. Section
\ref{sec:systematics} is dedicated to the impact of systematic effects
such as mismodeling the underlying physics or detector acceptance
effects. Numerical studies for one uni-angular and one triple-angular
distribution are provided in section \ref{sec:numerics}. In a series
of appendices we provide the orthonormal bases of angular functions
relevant to several rare $b$ decays.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ by means of
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

Integration over the non-angular variables yields
\begin{equation}
    \langle S_i\rangle
    \equiv \int \rmdx{\vec\nu} S_i(\nuvec)
    = \int \rmdx{\vec \nu} \left[\int_{\Omega} \rmdx{\vec \theta} P(\nuvec,\thvec) \dual{f}_i(\thvec) \right].
\end{equation}
The remainder of this section describes the method of moments, in
which we replace the analytical integration by Monte Carlo (MC)
estimates.  The central tenet of MC integration is the fact that the
expectation value $E_P[g]$ of some function $g(x)$ under the
probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int \rmdx{x} P(x) g(x),
\end{equation}
can be approximated by the consistent and unbiased
estimator $\est{E_P[g]}$~\cite[sec. 8.2]{James:2006zz}
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{N} \sum_{n=1}^N g(x^{(n)}) \,,\,    x^{(n)} \sim P
\end{equation}
due to the strong law of large numbers for $N \to \infty$, assuming
that the variates $x^{(n)}$, $n = 1, \dots, N$, are distributed as
$P$.
Throughout this letter we denote all MC estimators with a wide hat.\\

Application of \refeq{mc-id} then yields
\begin{equation}
    \langle S_i\rangle \to \widehat{\langle S_i\rangle} = \frac{1}{N} \sum_{n=1}^{N} \dual{f}_i(x^{(n)})\,.
\end{equation}
It is often of interest to obtain observables integrated over certain
bins of $\nuvec$. We define
\begin{align}
    \langle S_i\rangle_{\vec{a},\vec{b}}
    & \equiv \int_{\vec{a}}^{\vec{b}} \rmdx{\nuvec} S_i(\nuvec)\\
    & = \int_{\vec{a}}^{\vec{b}} \rmdx{\nuvec} \left[\int_{\Omega} \rmdx{\vec\theta} P(\nuvec,\thvec) \dual{f}_i(\thvec)  \right]\\
    & = \int \rmdx{\nuvec} \left[\int_{\Omega} \rmdx{\vec\theta}  P(\nuvec,\thvec) \dual{f}_i(\thvec)
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,
        \right],
\end{align}
where the argument of the indicator function $ \mathbf{1}(\vec{a} \le
\nuvec \le \vec{b})$ is to be interpreted componentwise.
Application of \refeq{mc-id} immediately yields
\begin{equation}
    \label{eq:bin-importance}
    \widehat{\langle S_i\rangle_{\vec{a},\vec{b}}}
    = \frac{1}{N} \sum_{n=1}^{N} \dual{f}_i(x^{(n)})         \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
For notational simplicity, let us forget about the $\nuvec$
integration and consider only $S_i$. In the limit $N \to \infty$, the
central limit theorem (CLT) implies that the random vector
\begin{equation}
  \label{eq:angular-obs-vec}
  \vecest{S} \equiv (\est{S_0}, \dots, \est{S_i}\,,
  \dots)
\end{equation}
follows a multivariate Gaussian distribution $\mathcal{N}(\vec{S},
\widehat{\Sigma})$ centered on the true value $\vec{S}$ with the
covariance $\widehat{\Sigma}_{ij}$ estimated as
\begin{equation}
    \Sigma_{ij} \equiv \cov[S_i,S_j] \to \est{\Sigma_{ij}} \equiv \est{\cov}[{S}_i, {S}_j]
        = \frac{1}{N - 1} \sum_{n=1}^{N} \Big[\dual{f}_i\big(x^{(n)}\big) - \est{S_i}\Big]\,\Big[\dual{f}_j\big(x^{(n)}\big) - \widehat{S_j}\Big]\,.
\end{equation}
In our physics applications, the parameter space is compact and each
$\dual{f}_i$ is bounded. Hence the requisites for the most basic
version of the CLT to hold --- finite mean and covariance of
$\dual{f}_i$ --- are automatically satisfied.
In our numerical analysis the sample covariance rapidly converges towards the true covariance
matrix; see also \refsec{numerics}.\\

Compared to the usual maximum-likelihood approach, we find for the method of moments:
\begin{enumerate}
  \item The angular observable ${S_i}$ can be determined
  independently of any other observable ${S_j}$. It is therefore
  much more robust to physics assumptions needed to define the full
  likelihood. In particular, this means one does not have to be
  specific regarding the form of new-physics contributions; in fact,
  one does not even need to be able to explicitly formulate the
  likelihood at all.
\item It is superior for a small number of samples $N$. Likelihood
  fits tend to be numerically unstable if lots of parameters need to
  be estimated from sparse data. This is more severe if the mode of
  the likelihood is near the boundary of the physically allowed
  region \cite{lehmann1998}. For some of these decays of interest, there are only $\order{100}$ events
  per bin on record.

 % \fred{Is there some LHCb note we could cite?}\danny{Ulrik suggests citing some
 % statistical paper}
 % For some of these decays of interest, there are only $\order{10}$--$\order{100}$ events
 % per bin on record.\danny{What to cite? 1403.8004 only states total signal yield!}
\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
  smaller than those from the method of moments.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

In several interesting processes we might only have an approximate
result for $P$.  In this section, we focus on one particular class of
mismodelling of the signal PDF: the angular-momentum cutoff in partial
wave expansions. This mismodelling potentially affects a large number
of decays and scattering processes. For the sake of clarity we take
the interesting class of four-body decays $B\to P_1 P_2 \ell_1 \ell_2$
as an example\footnote{This includes the rare $b\to s$ mediated $B$ decay $B
  \to K\pi\ell^+\ell^-$, and the $V_{ub}$ suppressed decay $B\to
  \pi\pi\ell^+\bar\nu_\ell$. For both examples the PDF $P$ is known in
  the small-width approximation and when assuming a pure $P$-wave,
  resonant final state. An extension to $S-P$ interference has been
  studied for $B\to K\pi\ell^+\ell^-$
  \cite{Blake:2012mb,Becirevic:2011bp}, and for $\bar{B}\to
  \pi\pi\ell^-\bar\nu_\ell$ \cite{Faller:2013dwa}. For a first study
  of $S$, $P$
  and $D$ interference, see \cite{Das:2014sra}
}.\\

@@

We wish to emphasize, however, that detector acceptance effects systematically affect the expansion for any basis
of angular functions, including the one suggested in this section.
Nevertheless, the expansion in terms of Legendre polynomials as suggested above provides means to cope with these effects,
as we discuss in the following subsection.

@@

\add{Apparently we now face a circular dependence. On the one hand,
the estimators $\vecest{Q}$, and thus also the estimator $\vecest{S}$
are proportional to $R$, the ratio of detected events over occuring events.
On the other hand, $R$ depends by construction (see \refeq{def-R})
on $P_\text{t}(x_t)$, and thus on the true value of the angular observables $\vec{S}$.
We solve this problem by observing that the MC estimators $\vecest{S}$ need to fulfill
a self-consistency condition, which reads
\begin{equation}
    \est{S}_0 = \frac{1}{2}\,,\qquad\text{for a uniangular distribution}.
\end{equation}
(For the process-dependent conditions in the multiangular case see \refeq{recipe:lambdabtolambdall} and
\refeq{recipe:btokstarll}.)
This self-consistency condition is tightly related to the determination of
the branching ratio of the underlying decay, and we therefore suggest to
carry out a combined analysis for the determination of the branching ratio
and the extraction of the angular observables.\\
Moreover, in the applications to $B$ decays only ratios and similar $R$-independent combinations
of the angular observables are of interest.
}\\

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S5_200.png}
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S7_200.png}
        \caption{Pull distribution for the angular observables $S_5$ (left) and $S_7$ (right), extracted from $2\cdot 10^5$ studies of
        $200$ simulated events each for the decay $B\to K^*\ell^+\ell^-$. The red curve represents a fit to a gaussian distribution.}
        \label{fig:pulls}
\end{figure}

>>>>

\begin{abstract}
  Our goal is to obtain a complete set of angular observables
  arising in a generic multi-body process without the need to carry
  out a likelihood fit of the angular distribution to the measured
  events. Instead, we apply the method of moments that relies
  both on the orthogonality of angular functions and the estimation of
  integrals by Monte Carlo techniques. The main advantage of
  this method is that the joint distribution of all observables
  given the data can be easily extracted, even for very few
  events. Our main contribution is to show explicitly how systematic
  uncertainties from detector-resolution and acceptance effects can
  be incorporated into a general algorithm. Finally, we present the
  necessary process-dependent formulae needed for direct
  implementation of the method for several rare decays of
  interest.
\end{abstract}

@@

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to arbitrary decay or scattering
processes \fred{Not really arbitrary, it only works with orthogonal
  basis functions}. We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

    \item \add{Likelihood fits  can be unstable in case the underlying
        physical model is only partially known. This can lead to
        overestimating the number of physical parameters, and consequently
        inhibits the convergence of the fits. As an example, this type
        of problem occured in toy studies of the decay $B\to K^*(\to K\pi)\ell^+\ell^-$
        as reported in \cite{Egede:1142152}. It was subsequently solved
        when a missing symmetry relation between the angular observables
        was found and applied, thereby reducing the number of fit parameters.
    }
    \item Mismodeling the underlying physics model can result in systematic bias
        in likelihood fits. We will
        show that the method of moments is insensitive to a certain type of mismodeling;
        i.e. introducing a cutoff in a partial-wave expansion of the signal PDF.

@@

\fred{For us, the major advantage is that the experiment can obtain all angular observables at once, and if there are more, one doesn't need more data. I'm not sure this becomes clear from the text, point 1.}

We continue with basic definitions that pertain to angular observables, and our
results in the subsequent sections. Let $\thvec$ denote the set of all
angles, and let $\nuvec$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $S_i$ as a coefficient in the probability density
function (PDF), $P(\nuvec, \thvec)$, of the process by
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) \equiv \sum_i S_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$\{f_i(\thvec)\}$. We assume there exists a dual basis of functions
$\{\dual{f}_i(\thvec)\}$ such that the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega \rmdx{\vec{\theta}} \dual{f}_i(\thvec) f_j(\thvec)  = \delta_{ij}
\end{equation}
hold with $\Omega$ representing the full angular phase space relevant
to the process.  For particle decays, $P$ is generally expressed in
terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\rmdx{^{2}\Gamma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\rmdx{^{2}\sigma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process. This change of normalization is equivalent to a linear rescaling of
the angular observables $\lbrace S_i\rbrace$; thus ratios or similar suitable combinations
of the angular observables are not affected by this change of normalization.\\

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. It turns out to be much more convenient to work
in the basis of Legendre polynomials but the statistical properties
are equivalent for both approaches \fred{I'm confused. What are the two approaches? Do we consider Legendre polynomials as an orthonormal basis or not?} provided that one replace $f_i \to
\dual{f}_i$ appropriately. Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:btokstarll}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

In the remainder of this letter we discuss how to obtain an angular
observable $S_i(\nuvec)$ in an experimental setup where each recorded
event is (approximately) distributed according to $P$.  We establish
the statistical basics in section \ref{sec:sample-based-det}. Section
\ref{sec:systematics} is dedicated to the impact of systematic effects
such as mismodeling the underlying physics or detector acceptance
effects. Numerical studies for one uni-angular and one triple-angular
distribution are provided in section \ref{sec:numerics}.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

Integration over the non-angular variables yields
\begin{equation}
    \langle S_i\rangle
    \equiv \int \rmdx{\vec\nu} S_i(\nuvec)
    = \int \rmdx{\vec \nu} \left[\int_{\Omega} \rmdx{\vec \theta} P(\nuvec,\thvec) \dual{f}_i(\thvec) \right].
\end{equation}
\fred{Why is there $\Omega$ but nothing for $\nuvec$?}
The remainder of this section describes the method of moments, in
which we replace the analytical integration by Monte Carlo (MC)
estimates.  The central tenet of MC integration is the fact that the
expectation value $E_P[g]$ of some function $g(x)$ under the
probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int \rmdx{x} P(x) g(x),
\end{equation}
can be approximated by the consistent and unbiased
estimator $\est{E_P[g]}$~\cite[sec. 8.2]{James:2006zz}
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{N} \sum_{n=1}^N g(x^{(n)}) \,,\,    x^{(n)} \sim P
\end{equation}
due to the strong law of large numbers for $N \to \infty$, assuming
that the variates $x^{(n)}$, $n = 1, \dots, N$, are distributed as
$P$.
Throughout this letter we denote all MC estimators with a wide hat.\\

Application of \refeq{mc-id} then yields
\begin{equation}
    \langle S_i\rangle \to \widehat{\langle S_i\rangle} = \frac{1}{N} \sum_{n=1}^{N} \dual{f}_i(x^{(n)})\,.
\end{equation}
It is often of interest to obtain observables integrated over certain
bins of $\nuvec$. We define
\begin{align}
    \langle S_i\rangle_{\vec{a},\vec{b}}
    & \equiv \int_{\vec{a}}^{\vec{b}} \rmdx{\nuvec} S_i(\nuvec)\\
    & = \int_{\vec{a}}^{\vec{b}} \rmdx{\nuvec} \left[\int_{\Omega} \rmdx{\vec\theta} P(\nuvec,\thvec) \dual{f}_i(\thvec)  \right]\\
    & = \int \rmdx{\nuvec} \left[\int_{\Omega} \rmdx{\vec\theta}  P(\nuvec,\thvec) \dual{f}_i(\thvec)
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,
        \right],
\end{align}
where the argument of the indicator function $ \mathbf{1}(\vec{a} \le
\nuvec \le \vec{b})$ is to be interpreted componentwise.
Application of \refeq{mc-id} immediately yields
\begin{equation}
    \label{eq:bin-importance}
    \widehat{\langle S_i\rangle_{\vec{a},\vec{b}}}
    = \frac{1}{N} \sum_{n=1}^{N} \dual{f}_i(x^{(n)})         \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
For notational simplicity, let us forget about the $\nuvec$
integration and consider only $S_i$. In the limit $N \to \infty$, the
central limit theorem (CLT) implies that the random vector
\begin{equation}
  \label{eq:angular-obs-vec}
  \vecest{S} \equiv (\est{S_0}, \dots, \est{S_i}\,,
  \dots)
\end{equation}
follows a multivariate Gaussian distribution $\mathcal{N}(\vec{S},
\Sigma)$ centered on the true value $\vec{S}$ with the
covariance $\Sigma_{ij}$ estimated as
\begin{equation}
    \Sigma_{ij} \equiv \cov[S_i,S_j] \to \est{\Sigma_{ij}} \equiv \est{\cov}[{S}_i, {S}_j]
        = \frac{1}{N - 1} \sum_{n=1}^{N} \Big[\dual{f}_i\big(x^{(n)}\big) - \est{S_i}\Big]\,\Big[\dual{f}_j\big(x^{(n)}\big) - \widehat{S_j}\Big]\,.
\end{equation}
In our physics applications, the parameter space is compact and each
$\dual{f}_i$ is bounded. Hence the requisites for the most basic
version of the CLT to hold --- finite mean and covariance of
$\dual{f}_i$ --- are automatically satisfied.
In our numerical analysis the sample covariance rapidly converges towards the true covariance
matrix; see also \refsec{numerics}.\\

Compared to the usual maximum-likelihood approach, we find for the method of moments:
\begin{enumerate}
  \item The angular observable ${S_i}$ can be determined
  independently of any other observable ${S_j}$. It is therefore
  much more robust to physics assumptions needed to define the full
  likelihood. In particular, this means one does not have to be
  specific regarding the form of new-physics contributions; in fact,
  one does not even need to be able to explicitly formulate the
  likelihood at all.
\item It is superior for a small number of samples $N$. Likelihood
  fits tend to be numerically unstable if lots of parameters need to
  be estimated from sparse data. This is more severe if the mode of
  the likelihood is near the boundary of the physically allowed
  region \cite{lehmann1998}. For some of these decays of interest, there are only $\order{100}$ events
  recorded per bin.

\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
  smaller than those from the method of moments  \fred{Is this true, Marcin?}.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

In several interesting processes we might only have an approximate
result for $P$.  In this section, we focus on one particular class of
mismodeling of the signal PDF: the angular-momentum cutoff in partial-wave expansions. This mismodeling potentially affects a large number
of decays and scattering processes. For the sake of clarity we take
the interesting class of four-body decays $B\to P_1 P_2 \ell_1 \ell_2$
as an example\footnote{This includes the rare $b\to s$ mediated $B$ decay $B
  \to K\pi\ell^+\ell^-$, and the $V_{ub}$ suppressed decay $B\to
  \pi\pi\ell^+\bar\nu_\ell$. For both examples the PDF $P$ is known in
  the small-width approximation and when assuming a pure $P$-wave
  resonant final state. An extension to $S-P$ interference has been
  studied for $B\to K\pi\ell^+\ell^-$
  \cite{Blake:2012mb,Becirevic:2011bp}, and for $\bar{B}\to
  \pi\pi\ell^-\bar\nu_\ell$ \cite{Faller:2013dwa}. For a first study
  of $S$, $P$
  and $D$ interference, see \cite{Das:2014sra}
}.\\

@@

We wish to emphasize that detector acceptance effects systematically affect the expansion for any basis
of angular functions, including the one suggested in this section.
Nevertheless, the expansion in terms of Legendre polynomials as suggested above provides means to cope with these effects,
as we discuss in the following subsection.

@@

\add{Apparently we now face a circular dependence. On the one hand,
the estimators $\vecest{Q}$ and thus also $\vecest{S}$
are proportional to $R$, the ratio of detected events over occuring events.
On the other hand, $R$ depends by construction (see \refeq{def-R})
on $P_\text{t}(x_t)$, and thus on the true value of the angular observables $\vec{S}$.
We solve this problem by observing that the MC estimators $\vecest{S}$ need to fulfill
the self-consistency condition
\begin{equation}
    \est{S}_0 = \frac{1}{2}\,,\qquad\text{for a uniangular distribution}.
\end{equation}
(For the process-dependent conditions in the multiangular case see \refeq{recipe:lambdabtolambdall} and
\refeq{recipe:btokstarll}.)
This self-consistency condition is tightly related to the determination of
the branching ratio of the underlying decay, and we therefore suggest to
carry out a combined analysis for the determination of the branching ratio
and the extraction of the angular observables \fred{Marchin, Nicola: Do you do a combined analysis for Moriond?}.\\
Moreover, in the applications to $B$ decays only ratios and similar $R$-independent combinations
of the angular observables are of interest.
}\\

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S5_200.png}
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S7_200.png}
        \caption{Pull distribution for the angular observables $S_5$ (left) and $S_7$ (right), extracted from $2\cdot 10^5$ studies of
        $200$ simulated events each for the decay $B\to K^*\ell^+\ell^-$. The red curve represents a fit to a Gaussian distribution.}
        \label{fig:pulls}
\end{figure}
